Metric Spaces with Expensive Distances

### 2018-11-29 paper/draft.tex
paper

\newcommand{\N}{\mathbb{N}}
\newcommand{\Q}{\mathbb{Q}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\Z}{\mathbb{Z}}
\renewcommand{\S}{\mathbb{S}}
\newcommand{\dappr}{\tilde{d}}
%\newcommand{\Wlog}{W.l.o.g.\ }
%\newcommand{\wlog}{w.l.o.g.\ }
\newcommand{\eps}{\epsilon}
\newcommand{\orient}{O}
\newcommand{\cpro}{\Pi}
\newcommand{\fib}{\mathrm{Fib}}

\newcommand{\parent}{\mathrm{par}}
\newcommand{\point}{p}
\newcommand{\level}{\ell}
\newcommand{\diam}{\mathrm{diam}}
\newcommand{\pointset}{P}
\newcommand{\distspace}{\mathcal{M}}
\newcommand{\dist}{\delta}
\newcommand{\adist}{\gamma}
\newcommand{\complexity}{C_{\dist}}
\newcommand{\doublingdimension}{\Delta}

@@

\title{Geometric Filters for Metric Spaces}
\author{Michael Kerber}{TU Graz}{kerber@tugraz.at}{}{}
\author{Arnur Nigmetov}{TU Graz}{nigmetov@tugraz.at}{}{}
\authorrunning{M. Kerber and A. Nigmetov}

@@

@@

\paragraph{Motivation.}

In this paper we look at algorithms in metric spaces,
in which computing a single distance between points can be very costly.
More specifically,
we consider the problem of finding an approximate nearest neighbor
of some point $q$ among the given $n$ points, and the problem of finding a $t$-spanner,
i.e., a graph that approximates the metric on the given set of points,
and we not so much interested in reducing the complexity of the algorithm,
but rather in reducing the number of distances computed.
The motivating example for us was Topological Data Analysis, where computation
of bottleneck and Wasserstein distances between two persistence diagrams
can be done in time polynomial in the number of points of the input diagrams.

We believe that the scenario when the input consists of a relatively small number of points, 
e.g., of order $10^2-10^3$, but the complexity of computing a distance is rather large,
is plausible (for the persistence diagrams, the size of 0-dimensional ones can be huge,
because in dimension 0 there are especially efficient algorithms that can handle much
larger inputs, so it is possible to produce diagrams with millions of  points in each).
In such a situation, the natural approach would be to avoid computing the
distance between two points if we do not need it,

We demonstrate that the number of distance computations can be further
reduced by a careful re-implementation of the underlying geometric algorithms.
Our improvement is based on the following simple observation: many geometric
constructions (such as cover tree and WSPD constructions) require distance
computations only in the form of the following predicate: 
``Is $\dist(p_i,p_j)\leq t$'', where $t$ is a real value.
We show that using the information of the already computed distances and
the triangle inequality, we can answer the above predicate in many cases
without computing the actual distance.

We describe our idea with an example. In Figure~\ref{fig:1st_example},
we see a weighted graph on $P$ whose edges are the distances that have
been computed so far; note that $\dist(p_1,p_2)$ is unknown. 
The shortest path from $p_1$ to $p_2$ has length $9$, which clearly
constitutes an upper bound on the distance by triangle inequality.
However, we can also infer that $\dist(p_1,p_2)\geq 3$:
otherwise, the path from $p_3$ to $p_4$ via $p_1$ and $p_2$
would be shorter than the edge $(p_3,p_4)$, again contradicting
triangle inequality.
Hence, the question whether $\dist(p_1,p_2)\leq t$ can be answered positively
for $t>=9$ and negatively for $t<3$. Only if $t\in[3,9)$, the actual
distance computation is necessary. This approach is reminiscent to the usage
of \emph{numerical filters} in geometric computing in the CGAL library,
where geometric predicates in exact number types are first evaluated using 
certified floating-point arithmetic, and exact computation is only
invoked if the numerical evaluation did not arrive at a decisive result
(e.g.~\cite{bbp-interval,kerber-phd}).

Given a set $\pointset:=\{p_1,\ldots,p_n\}$ of $n$ points
in a metric space $(\distspace,\dist)$, 
consider the following standard operations:

\begin{description}
\item[Nearest Neighbor] Given a point $q\in\distspace$,
find $p_i\in\pointset$ such that, for all $j=1,\ldots,n$,
\[d(q,p_i)\leq d(q,p_j)\]

\item[Metric approximation] Given $\eps>0$, compute a metric $\adist$
such that for all $1\leq i,j\leq n$,
%
\[\dist(p_i,p_j)\leq \adist(p_i,p_j) \leq (1+\eps) \dist(p_i,p_j).\]

By ``computing'' $\adist$, we mean to compute a $n\times n$-matrix $E$ 
such that $E_{ij}=\adist(p_i,p_j)$.
\end{description}

Assume that an algorithm is given to compute $\dist(\cdot,\cdot)$
for any two points in $\distspace$ in worst-case complexity $\complexity$.
Then trivially, the above problems can be solved (exactly) in
$O(n\complexity)$ and $O(n^2\complexity)$ time, respectively.
% For low-dimensional\footnote{we will make this notion precise
% later in the text} metric spaces, we can improve these bounds:
% Using cover trees~\cite{cover-trees}, we can preprocess the point set
% in $O(\complexity n\log n)$ time to answer nearest neighbor queries
% in $O(\complexity \log n)$ time. Combining cover trees with well-separated
% pair decompositions (WSPD), we obtain an $\eps$-spanner for the graph
% induced by $\dist$ in $O(\complexity n\log n)$ time and
% of linear size (assuming, for simplicity, that $\eps$ is a constant).
% Out of this spanner,
% we can construct an approximate metric
% with an all-pair shortest path computation,
% resulting in a total running time of
% \[O(\complexity n\log n+ n^2\log n)\]
% using Johnson's algorithm~\cite{johnson}.

@@

\paragraph{Contribution.}

%
We demonstrate that the number of distance computations can be further
reduced by a careful re-implementation of the underlying geometric algorithms.
Our improvement is based on the following simple observation: many geometric
constructions (such as cover tree and WSPD constructions) require distance
computations only in the form of the following predicate: 
``Is $\dist(p_i,p_j)\leq t$'', where $t$ is a real value.
We show that using the information of the already computed distances and
the triangle inequality, we can answer the above predicate in many cases
without computing the actual distance.

We describe our idea with an example. In Figure~\ref{fig:1st_example},
we see a weighted graph on $P$ whose edges are the distances that have
been computed so far; note that $\dist(p_1,p_2)$ is unknown. 
The shortest path from $p_1$ to $p_2$ has length $9$, which clearly
constitutes an upper bound on the distance by triangle inequality.
However, we can also infer that $\dist(p_1,p_2)\geq 3$:
otherwise, the path from $p_3$ to $p_4$ via $p_1$ and $p_2$
would be shorter than the edge $(p_3,p_4)$, again contradicting
triangle inequality.
Hence, the question whether $\dist(p_1,p_2)\leq t$ can be answered positively
for $t>=9$ and negatively for $t<3$. Only if $t\in[3,9)$, the actual
distance computation is necessary. This approach is reminiscent to the usage
of \emph{numerical filters} in geometric computing in the CGAL library,
where geometric predicates in exact number types are first evaluated using 
certified floating-point arithmetic, and exact computation is only
invoked if the numerical evaluation did not arrive at a decisive result
(e.g.~\cite{bbp-interval,kerber-phd}).

\begin{figure}[h]

\centering
\includegraphics[width=6cm]{intro_example.eps}
\caption{An example of a partially revealed metric.}
\label{fig:1st_example}
\end{figure}

In our implementation, we maintain the best lower and upper bounds of all
pairwise distances in an $n\times n$ matrix of intervals. We show how to
update this matrix in $O(n^4)$ time when a new edge is added.
%We point out once more that a quadratic update time only pays off
%if distance computations are expensive, say $\complexity=\Omega(n^2)$.

We perform an extensive experimental evaluation of our idea.
We mostly concentrate on point clouds in Euclidean space~--
of course, the approach does not yield any performance benefits in that
situation because distance computations are cheap; however, the Euclidean
case provides a simple setup for experiments.
We investigate how the dimensionality, the clustering density, 
the approximation quality, and noise 
in the point cloud affect the number of distance computations.
Our result show that $\ldots$

We also provide some experimental results for the case of persistence diagrams
which motivated this line of research. We point out that the space of
persistence diagrams fails to be ``low-dimensional''. To be precise,
its doubling dimension is $\infty$, and a finite set of $n$ diagram
can have the worst-case doublng dimension $\log n$. This worst-case
behavior is also the expected one if the diagrams are chosen randomly
(in a sense made precise in Section~\ref{sec:metric_spaces_top_summaries}). 
However, we demonstrate that
for natural collections of persistence diagrams, a substantial fraction
of distance computations can be avoided. In detail, $\ldots$

@@

\section{Notation and Preliminary Notions}

@@

We shall need the following lemma.

\begin{lemma}
\label{lem:packing_lemma_dd}
Let $(X,\dist)$ be a metric space of doubling dimension $d$, and let $P$ be a subset of a ball $B(x,R)$ in $X$ such that the distance between any two distinct points of $P$ is at least $r$.
Then $P$ is finite and cardinality of $P$ satisfies the inequality $|P| < 2^{md}$, where $m = 1 + \lceil \log (R/r) \rceil $. 
\end{lemma}

\begin{proof}
We can cover $B(x,R)$ with $2^d$ ball of radius $R/2$, each of these balls we can cover with $2^d$
balls of radius $R/4$, etc. Repeating this process $\lceil \log \frac{R}{r/2} \rceil = m $ times, 
we cover
$B(x, R)$ with $2^{md}$ balls of radius at most $r/2$. Since a ball of radius $r/2$ can contain at most one point from $P$, the claim follows.
\end{proof}

@@

It is known that asymptotically optimal spanner can be constructed by the following greedy algorithm: 
start with an empty graph on vertices $p_i$,
sort all edges $\{p_i, p_j\}$ by the length $\dist(p_i, p_j)$ in ascending order and keep adding them to the graph
until it becomes a $1+\eps$-spanner.

 In our setting this is the most expensive approach, because it requires
 computing all pairwise distances first. We propose an alternative approach (or rather a family of algorithms), which we call
 \textit{blind} spanner, meaning that the algorithm that builds a spanner does not have access to $\dist(p_i, p_j)$
 before it decides to add the edge $(p_i, p_j)$.
 The algorithm also maintains lower and upper bounds
 for all pairwise distances, which are updated after a new edge is added. and the stopping criterion
 is based on these bounds, namely, if $a_{i, j}$ and $b_{i,j}$ denote lower and upper bounds
 for the distance between $p_i$ and $p_j$, respectively, then the algorithm stops, if $b_{i,j} / a_{i, j} \leq 1 + \eps$
 for all $i \neq j$.

\begin{algorithmic}
\label{alg:blind_spanner}
\Function{BlindSpanner}{$P, \eps$}
    \State {$E \gets \emptyset$}
    \State {$a_{i,j} \gets 0$ for all $1 \leq i,j \leq n$}
    \State {$b_{i,j} \gets \infty$ for all $1 \leq i,j \leq n, i \neq j$}
    \While {$\exists i \neq j : b_{i,j} / a_{i,j} > 1 + \eps$}
    \State {$(i,j) \gets GetEdgeToAdd$}
    \State {$v \gets \dist(p_i, pj)$}
    \State {Add weighted edge $(p_i, p_j, v)$ to $E$}
    \State {$UpdateBounds(i, j, v)$}
    \EndWhile
\EndFunction
\end{algorithmic}

In this pseudocode we adopt the convention that a positive number divided by 0 is $\infty$
and $\infty$ is larger than any real number,
thus making the predicate in the while loop well-defined.  The procedure $UpdateBounds$ is desribed in the next subsection,
now we only mention that it updates all upper and lower bounds based on triangle inequality involving the edge $(p_i, p_j)$,
and the update is monotonic: no upper bound can increase, no lower bound can decrease, and, after the distance
between $p_i$ and $p_j$ is calculated, both $a_{i,j}$ and  $b_{i,j}$ are set to be $\dist(p_i, p_j)$.
The most important part of this algorithm is the way of choosing the next edge to be added, and
there are several possibilities, which give very different results.

@@

 \paragraph{Maintaining Lower and Upper Bounds}

@@

\paragraph{Experimental Results}
At the moment we cannot provide any theoretical guarantees for the heuristics that we propose to compute blind spanners,
and it can happen that it will yield the complete graph as a spanner.
We run extensive experiments on the points sampled from the low-dimensional Euclidean space to investigate
experimentally the performance of these heuristics. Clearly, this has no practical meaning for $\R^d$, but
this space is well-suited for controlled experiments.

\begin{figure}[ht]
    \label{fig:spanner-sparseness}
    \includegraphics[width=0.7\textwidth]{edges_to_points_ratio_dim_2_normal_points.png}
    \caption{Number of edges in blind spanners generated by different variants of the blind
    algorithm. Greedy non-blind algorithm is included for comparison. The plot is for normally distributed points
    in dimension 2, $\eps = 0.1$.}
\end{figure}

\begin{figure}[ht]
    \label{fig:blind-greedy-only}
    \includegraphics[width=0.7\textwidth]{blind-greedy-dims.png}
    \caption{}
\end{figure}

We tested the algorithm for $\eps \in \Set{0.01, 0.1, 0.2, 0.5}$ on the following sets of points in dimensions $d = 2,3,4,5$:
\begin{enumerate}
    \item In the \textbf{uniform} test set points are sampled uniformly at
        random from the unit cube in $\mathbb{R}^d$.
    \item In the \textbf{normal} test set points are sampled from the standard
        normal distribution in $\mathbb{R}^d$.
    \item In the \textbf{clustered} test set we first sample cluster centers uniformly 
        at random from $[0,10000]^d$, and then we add normally distributed noise around
        each of the centers. The number of clusters is chosen so that each cluster
        contains 50 points.
    \item The test set \textbf{exp} consists of points of the form $(2^{\xi_1}, \dots, 2^{xi_d})$,
        where $\xi_i$'s are i.i.d. random variables with uniform distribution on $[1,25]$.
\end{enumerate}
In all experiments the algorithms that we tested compared in the same way,
so we only present results for the \textbf{uniform} point set in dimension 2.
%Let $e$ denote the number of edges of a spanner on $n$ vertices;
%in the plots we visualize the ratios $e / { n \choose 2}$ and $e / n$.
%We refer to the first ratio as \textit{density} and to the second one
%as \textit{growth}. Note that

The results are summarized in the plot \ref{fig:spanner-sparseness}.
The first plot shows the number of edges of the spanner.
We can see that, even though none of the blind spanners can produce
spanners of the same quality (i.e., sparseness) as the standard greedy algorithm,
\textsc{BlindGreedy} and all variants of \textsc{BlindRandom}
perform significantly better than both variants of \textsc{BlindQuasiSorted}.
it is important to remember that for all blind spanners the number of edges
is equal to the number of distances that were actually computed, while
for non-blind greedy spanner the latter number is always $n \choose 2$.
The second plot shows the ratio of the number of edges to the number of points.
The ideal behavior is demonstrated by the non-blind greedy spanner,
for which this ratio stays practically constant, confirming the linear growth.
None of the blind algorithms seems to have this property, but among them
the blind greedy spanner is the best one. If we assume that the number of edges
is proportional to $n^\alpha$, then we can try to estimate $\alpha$ by 
linear regression.  We give in the table \ref{tbl:regr_coeff_spanner}
the maximal exponents $\alpha$ for each of the algorithms,
and it is worth mentioning that even for the quasi-sorted shaker variant 
the exponent seems to be bounded away from 2.

As for different variants of the \textsc{BlindRandomBadRatio} algorithm,
we note that their performance is almost the same, and the algorithm works
significantly better than \textsc{QuasiSorted} variants, but obviously
worse than the blind greedy variant. There is a consistent, though small,
difference between the variants that do not force lower bounds first and
the other two variants of the BlindRandomBadRatio.

>>>>

\newcommand{\N}{\mathbb{N}}
\newcommand{\Q}{\mathbb{Q}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\Z}{\mathbb{Z}}
\renewcommand{\S}{\mathbb{S}}
\newcommand{\dappr}{\tilde{d}}
%\newcommand{\Wlog}{W.l.o.g.\ }
%\newcommand{\wlog}{w.l.o.g.\ }
\newcommand{\eps}{\varepsilon}
\newcommand{\orient}{O}
\newcommand{\cpro}{\Pi}
\newcommand{\fib}{\mathrm{Fib}}

\newcommand{\metricspace}{\mathcal{M}}
\newcommand{\approxmetric}{\gamma}
\newcommand{\parent}{\mathrm{par}}
\newcommand{\point}{p}
\newcommand{\level}{\ell}
\newcommand{\diam}{\mathrm{diam}}
\newcommand{\pointset}{P}
\newcommand{\distspace}{\mathcal{M}}
\newcommand{\dist}{\delta}
\newcommand{\adist}{\gamma}
\newcommand{\complexity}{C_{\dist}}
\newcommand{\doublingdimension}{\Delta}

@@

\title{Metric Spaces with Expensive Distances}
\author{Michael Kerber}{TU Graz}{kerber@tugraz.at}{}{}
\author{Arnur Nigmetov}{TU Graz}{nigmetov@tugraz.at}{}{}
\authorrunning{M. Kerber and A. Nigmetov}

@@

\begin{abstract}

In algorithms for finite metric spaces, it is common
to assume that the distance between two points
can be computed in constant time, and complexity
bounds are expressed only in terms of the number
of points of the metric space.
We introduce a different model where we assume
that the computation of a single distance is
an expensive operation and consequently, the goal is
to minimize the number of such distance queries.
This model is motivated by metric spaces
that appear in the context of topological data analysis.

We consider two standard operations on metric spaces,
namely the construction of an $\epsilon$-spanner
and the computation of an approximate nearest
neighbor for a given query point.
In both cases, we partially explore the metric space
through distance queries and infer lower and upper bounds
for yet unexplored distances through triangle inequality.
For spanners, we evaluate several exploration strategies
through extensive experimental evaluation.
For approximate nearest neighbors, we prove that our
strategy returns an approximate nearest neighbor
after a logarithmic number of distance queries.
\end{abstract}

@@

\subsection{Motivation.}

In this paper we look at algorithms in metric spaces,
in which computing a single distance between points can be very costly.
The motivating example for us was Topological Data Analysis, where computation
of bottleneck and Wasserstein distances between two persistence diagrams
can be done in time polynomial in the number of points of the input diagrams.

Given a set $\pointset:=\{p_1,\ldots,p_n\}$ of $n$ points
in a metric space $(\metricspace,\dist)$, 
consider the following standard operations:

\begin{description}
\item[Nearest Neighbor] Given a point $q\in\metricspace$,
find $p_i\in\pointset$ such that, for all $j=1,\ldots,n$,
\[d(q,p_i)\leq d(q,p_j)\]

\item[Metric approximation] Given $\eps>0$, compute a metric $\approxmetric$
such that for all $1\leq i,j\leq n$,
%
\[\dist(p_i,p_j)\leq\approxmetric(p_i,p_j)\leq (1+\eps)\dist(p_i,p_j).\]

By ``computing'' $\approxmetric$, we mean to compute a $n\times n$-matrix $E$ 
such that $E_{ij}=\approxmetric(p_i,p_j)$.
\end{description}

% We believe that the scenario when the input consists of a relatively small number of points, 
% e.g., of order $10^2-10^3$, but the complexity of computing a distance is rather large,
% is plausible (for the persistence diagrams, the size of 0-dimensional ones can be huge,
% because in dimension 0 there are especially efficient algorithms that can handle much
% larger inputs, so it is possible to produce diagrams with millions of  points in each).
% In such a situation, the natural approach would be to avoid computing the
% distance between two points if we do not need it,

% We demonstrate that the number of distance computations can be further
% reduced by a careful re-implementation of the underlying geometric algorithms.
% Our improvement is based on the following simple observation: many geometric
% constructions (such as cover tree and WSPD constructions) require distance
% computations only in the form of the following predicate: 
% ``Is $\dist(p_i,p_j)\leq t$'', where $t$ is a real value.
% We show that using the information of the already computed distances and
% the triangle inequality, we can answer the above predicate in many cases
% without computing the actual distance.

% We describe our idea with an example. In Figure~\ref{fig:1st_example},
% we see a weighted graph on $P$ whose edges are the distances that have
% been computed so far; note that $\dist(p_1,p_2)$ is unknown. 
% The shortest path from $p_1$ to $p_2$ has length $9$, which clearly
% constitutes an upper bound on the distance by triangle inequality.https://v2.overleaf.com/project/5bf6fa26032afa24d94383a5
% However, we can also infer that $\dist(p_1,p_2)\geq 3$:
% otherwise, the path from $p_3$ to $p_4$ via $p_1$ and $p_2$
% would be shorter than the edge $(p_3,p_4)$, again contradicting
% triangle inequality.
% Hence, the question whether $\dist(p_1,p_2)\leq t$ can be answered positively
% for $t>=9$ and negatively for $t<3$. Only if $t\in[3,9)$, the actual
% distance computation is necessary. This approach is reminiscent to the usage
% of \emph{numerical filters} in geometric computing in the CGAL library,
% where geometric predicates in exact number types are first evaluated using 
% certified floating-point arithmetic, and exact computation is only
% invoked if the numerical evaluation did not arrive at a decisive result
% (e.g.~\cite{bbp-interval,kerber-phd}).

% Given a set $\pointset:=\{p_1,\ldots,p_n\}$ of $n$ points
% in a metric space $(\distspace,\dist)$, 
% consider the following standard operations:

% \begin{description}
% \item[Nearest Neighbor] Given a point $q\in\distspace$,
% find $p_i\in\pointset$ such that, for all $j=1,\ldots,n$,
% \[d(q,p_i)\leq d(q,p_j)\]

% \item[Metric approximation] Given $\eps>0$, compute a metric $\adist$
% such that for all $1\leq i,j\leq n$,
% %
% \[\dist(p_i,p_j)\leq \adist(p_i,p_j) \leq (1+\eps) \dist(p_i,p_j).\]

% By ``computing'' $\adist$, we mean to compute a $n\times n$-matrix $E$ 
% such that $E_{ij}=\adist(p_i,p_j)$.
% \end{description}

% Assume that an algorithm is given to compute $\dist(\cdot,\cdot)$
% for any two points in $\distspace$ in worst-case complexity $\complexity$.
% Then trivially, the above problems can be solved (exactly) in
% $O(n\complexity)$ and $O(n^2\complexity)$ time, respectively.
% For low-dimensional\footnote{we will make this notion precise
% later in the text} metric spaces, we can improve these bounds:
% Using cover trees~\cite{cover-trees}, we can preprocess the point set
% in $O(\complexity n\log n)$ time to answer nearest neighbor queries
% in $O(\complexity \log n)$ time. Combining cover trees with well-separated
% pair decompositions (WSPD), we obtain an $\eps$-spanner for the graph
% induced by $\dist$ in $O(\complexity n\log n)$ time and
% of linear size (assuming, for simplicity, that $\eps$ is a constant).
% Out of this spanner,
% we can construct an approximate metric
% with an all-pair shortest path computation,
% resulting in a total running time of
% \[O(\complexity n\log n+ n^2\log n)\]
% using Johnson's algorithm~\cite{johnson}.

@@

\subsection{Contribution.}
We consider the two well-known problems: finding a $1 + \eps$-spanner
and finding an approximate nearest neighbor.
%
% We demonstrate that the number of distance computations can be further
% reduced by a careful re-implementation of the underlying geometric algorithms.
% Our improvement is based on the following simple observation: many geometric
% constructions (such as cover tree and WSPD constructions) require distance
% computations only in the form of the following predicate: 
% ``Is $\dist(p_i,p_j)\leq t$'', where $t$ is a real value.
% We show that using the information of the already computed distances and
% the triangle inequality, we can answer the above predicate in many cases
% without computing the actual distance.

% We describe our idea with an example. In Figure~\ref{fig:1st_example},
% we see a weighted graph on $P$ whose edges are the distances that have
% been computed so far; note that $\dist(p_1,p_2)$ is unknown. 
% The shortest path from $p_1$ to $p_2$ has length $9$, which clearly
% constitutes an upper bound on the distance by triangle inequality.
% However, we can also infer that $\dist(p_1,p_2)\geq 3$:
% otherwise, the path from $p_3$ to $p_4$ via $p_1$ and $p_2$
% would be shorter than the edge $(p_3,p_4)$, again contradicting
% triangle inequality.
% Hence, the question whether $\dist(p_1,p_2)\leq t$ can be answered positively
% for $t>=9$ and negatively for $t<3$. Only if $t\in[3,9)$, the actual
% distance computation is necessary. This approach is reminiscent to the usage
% of \emph{numerical filters} in geometric computing in the CGAL library,
% where geometric predicates in exact number types are first evaluated using 
% certified floating-point arithmetic, and exact computation is only
% invoked if the numerical evaluation did not arrive at a decisive result
% (e.g.~\cite{bbp-interval,kerber-phd}).

% \begin{figure}[h]

% \centering
% \includegraphics[width=6cm]{intro_example.eps}
% \caption{An example of a partially revealed metric.}
% \label{fig:1st_example}
% \end{figure}

% In our implementation, we maintain the best lower and upper bounds of all
% pairwise distances in an $n\times n$ matrix of intervals. We show how to
% update this matrix in $O(n^2)$ time when a new edge is added.
% %We point out once more that a quadratic update time only pays off
% %if distance computations are expensive, say $\complexity=\Omega(n^2)$.

% We perform an extensive experimental evaluation of our idea.
% We mostly concentrate on point clouds in Euclidean space~--
% of course, the approach does not yield any performance benefits in that
% situation because distance computations are cheap; however, the Euclidean
% case provides a simple setup for experiments.
% We investigate how the dimensionality, the clustering density, 
% the approximation quality, and noise 
% in the point cloud affect the number of distance computations.
% Our result show that $\ldots$

% We also provide some experimental results for the case of persistence diagrams
% which motivated this line of research. We point out that the space of
% persistence diagrams fails to be ``low-dimensional''. To be precise,
% its doubling dimension is $\infty$, and a finite set of $n$ diagram
% can have the worst-case doublng dimension $\log n$. This worst-case
% behavior is also the expected one if the diagrams are chosen randomly
% (in a sense made precise in Section~\ref{sec:metric_spaces_top_summaries}). 
% However, we demonstrate that
% for natural collections of persistence diagrams, a substantial fraction
% of distance computations can be avoided. In detail, $\ldots$

@@

\section{Notation and Preliminary Notions}
Let $(X, \dist)$ be a finite metric space with $n$. Let $G$ be a graph with set of vertices $X$,
its edges have natural weighting $w(x,y) = \dist(x, y)$ which induces the graph metric $\dist_G$ on $X$.
Graph $X$ is called a $1 + \eps$-spanner of $P$, if $\dist_G$ is a good approximation of $\dist$:
$\dist(x,y) \leq \dist_G(x, y) \leq (1+\eps) \dist(x, y)$. Obviously, $G$ must have $\Omega(n)$ edges,
and we consider $G$ optimal, if it has $O(n)$ edges (for each edge of $G$ we must compute the corresponding
distance, if we want to use the spanner instead of the original metric, which dictates this
choice of optimality measure).

@@

We shall need the following lemma, which is just a reformulation of the well-known
packing lemma for doubling spaces (see \cite{smid_2009}, Sect. 2.2).

\begin{lemma}
\label{lem:packing_lemma}
 Let $(X,\dist)$ be a metric space of doubling dimension $d$, and let $P$ be a subset of a ball 
 $B(x,R)$ in $X$ such that the distance between any two distinct points of $P$ is at least $r$.
 Then $P$ is finite and its cardinality is bounded by some $C$ that depends only on $d$
 and the ratio $r/R$.
% Then $P$ is finite and cardinality of $P$ satisfies the inequality $|P| < 2^{md}$, where $m = 1 + \lceil \log (R/r) \rceil $.

% Let $(X,\dist)$ be a metric space of doubling dimension $d$, and let $P$ be a subset of a ball $B(x,R)$ in $X$ such that the distance between any two distinct points of $P$ is at least $r$.
% Then $P$ is finite and cardinality of $P$ satisfies the inequality $|P| < 2^{md}$, where $m = 1 + \lceil \log (R/r) \rceil $. 
\end{lemma}
\begin{proof}
We can cover $B(x,R)$ with $2^d$ ball of radius $R/2$, each of these balls we can cover with $2^d$
balls of radius $R/4$, etc. Repeating this process $m := \lceil \log \frac{R}{r/2} \rceil$ times, 
we cover
$B(x, R)$ with $2^{md}$ balls of radius at most $r/2$. Since a ball of radius $r/2$ can
contain at most one point from $P$, the claim follows.
\end{proof}
If the doubling dimension of $(X, \dist)$ is larger than $d$, then
there exists a ball of radius $R$ such that there are at least $2^d$ points with pairwise distance at least $R$. On the other hand,
the maximal distance between any two of these points is at most $2R$, and we have a lot of points with approximately the same distance
between them. If we want a high-quality spanner on these points, then we can expect that its size will be quadratic, not linear;
so the space of high doubling dimension cannot have optimal spanners. Thus the question of approximating
the non-doubling metric without computing many distances cannot be solved positively, and we should
restrict our attention to the case of doubling spaces.

@@

It is known that optimal spanner can be constructed by the following greedy algorithm: 
start with an empty graph on vertices $p_i$,
sort all edges $\{p_i, p_j\}$ by the length $\dist(p_i, p_j)$ in ascending order and keep adding them to the graph
until it becomes a $1+\eps$-spanner.

 In our setting this is the most expensive approach, because it requires
 computing all pairwise distances first. We propose an alternative approach (or rather a family of algorithms), which we call
 \textit{blind} spanner, meaning that the algorithm that builds a spanner does not have access to $\dist(p_i, p_j)$
 before it decides to add the edge $(p_i, p_j)$.
 The algorithm also maintains lower and upper bounds
 for all pairwise distances, which are updated after a new edge is added.
 These bounds are used in the stopping criterion, namely, if $a_{i, j}$ and $b_{i,j}$ denote lower and upper bounds
 for the distance between $p_i$ and $p_j$, respectively, 
 then the algorithm stops, if $b_{i,j} / a_{i, j} \leq 1 + \eps$
 for all $i \neq j$.

\begin{algorithmic}
\label{alg:blind_spanner}
\Function{BlindSpanner}{$P, \eps$}
    \State {$E \gets \emptyset$}
    \State {$a_{i,j} \gets 0$ for all $1 \leq i,j \leq n$}
    \State {$b_{i,j} \gets \infty$ for all $1 \leq i,j \leq n, i \neq j$}
    \While {$\exists i \neq j : b_{i,j} / a_{i,j} > 1 + \eps$}
    \State {$(i,j) \gets $} \Call{GetEdgeToAdd}{}
    \State {$v \gets \dist(p_i, pj)$}
    \State {Add weighted edge $(p_i, p_j, v)$ to $E$}
    \State \Call{UpdateBounds}{$i, j, v$}
    \EndWhile
\EndFunction
\end{algorithmic}

In this pseudocode we adopt the convention that a positive number divided by 0 is $\infty$
and $\infty$ is larger than any real number,
thus making the predicate in the while loop well-defined.  The procedure \textsc{UpdateBounds} is described in the next subsection,
now we only mention that it updates all upper and lower bounds based on triangle inequality involving the edge $(p_i, p_j)$,
and the update is monotonic: no upper bound can increase, no lower bound can decrease, and, after the distance
between $p_i$ and $p_j$ is calculated, both $a_{i,j}$ and  $b_{i,j}$ are set to be $\dist(p_i, p_j)$.
The most important part of this algorithm is the way of choosing the next edge to be added, and
there are several possibilities, which give very different results.

@@

 \subsection{Maintaining Lower and Upper Bounds}

@@

\subsection{Experimental Results}
At the moment we cannot provide any theoretical guarantees for the heuristics that we propose to compute blind spanners,
and it can happen that it will yield the complete graph as a spanner.
We run experiments on the points sampled from the low-dimensional Euclidean space to investigate
experimentally the performance of these heuristics. Clearly, this has no practical meaning for $\R^d$, but
this space is well-suited for controlled experiments.

\begin{figure}[ht]
    \label{fig:spanner-sparseness}
    \includegraphics[width=0.7\textwidth]{pics/edges_to_points_ratio_dim_2_normal_points.png}
    \caption{Number of edges in blind spanners generated by different variants of the blind
    algorithm. Greedy non-blind algorithm is included for comparison. The plot is for normally distributed points
    in dimension 2, $\eps = 0.1$.}
\end{figure}

\begin{figure}[ht]
    \label{fig:blind-greedy-only}
    \includegraphics[width=0.7\textwidth]{pics/blind-greedy-dims.png}
    \caption{}
\end{figure}

\begin{figure}[ht]
    \label{fig:blind-rbr-variants}
    \includegraphics[width=0.7\textwidth]{pics/random-bad-ratio-comparison.png}
    \caption{Comparison of the four variants of \textsc{BlindRandomBadRatio} algorithm.}
\end{figure}

We tested the algorithm for $\eps \in \Set{0.01, 0.1, 0.2, 0.5}$ on the following sets of points in dimensions $d = 2,3,4,5$:
\begin{enumerate}
    \item In the \textbf{uniform} test set points are sampled uniformly at
        random from the unit cube in $\mathbb{R}^d$.
    \item In the \textbf{normal} test set points are sampled from the standard
        normal distribution in $\mathbb{R}^d$.
    \item In the \textbf{clustered} test set we first sample cluster centers uniformly 
        at random from $[0,10000]^d$, and then we add normally distributed noise around
        each of the centers. The number of clusters is chosen so that each cluster
        contains 50 points.
    \item The test set \textbf{exp} consists of points of the form $(2^{\xi_1}, \dots, 2^{xi_d})$,
        where $\xi_i$'s are i.i.d. random variables with uniform distribution on $[1,25]$.
\end{enumerate}
In all experiments the algorithms that we tested compared in the same way,
so we only present results for the \textbf{uniform} point set in dimension 2.

The plot \ref{fig:spanner-sparseness}.
shows the number of edges of the spanner.
We can see that, even though none of the blind spanners can produce
spanners of the same quality (i.e., sparse) as the standard greedy algorithm,
\textsc{BlindGreedy} and all variants of \textsc{BlindRandom}
perform significantly better than both variants of \textsc{BlindQuasiSorted}.
it is important to remember that for all blind spanners the number of edges
is equal to the number of distances that were actually computed, while
for non-blind greedy spanner the latter number is always $n \choose 2$.
The second plot shows the ratio of the number of edges to the number of points.
The ideal behavior is demonstrated by the non-blind greedy spanner,
for which this ratio stays practically constant, confirming the linear growth.
None of the blind algorithms seems to have this property, but among them
the blind greedy spanner is the best one. If we assume that the number of edges
is proportional to $n^\alpha$, then we can try to estimate $\alpha$ by 
linear regression (after taking $\log$).  We give in the table \ref{tbl:regr_coeff_spanner}
the estimated exponents $\alpha$ for \textsc{BlindGreedy} and standard
greedy algorithms. Note that even for the greedy algorithm these estimated
exponents can be significantly larger than 1,
which is explained by the fact that the number of points
on which we computed is not large enough to clearly see
the linear dependency.

\begin{table}[]
\label{tbl:regr_coeff_spanner}
\begin{tabular}{|l|l|l|}
\hline
dimension & \textbf{Greedy (non-blind)} & \textbf{Blind greedy} \\ \hline
2         &         1.08                &  1.12                 \\ \hline
3         &         1.24                &  1.41                \\ \hline
4         &         1.42                &  1.77                 \\ \hline
\end{tabular}
\caption{Estimated exponents in the $|E|= C |V|^\alpha$ dependency of the number of edges
on the number of points. The data is for $\eps = 0.1$ and for uniform points.}
\end{table}

As for different variants of the \textsc{BlindRandomBadRatio} algorithm,
we note that their performance is almost the same, and the algorithm works
significantly better than \textsc{QuasiSorted} variants, but obviously
worse than the blind greedy variant. There is a consistent, though small,
difference between the variants that do not force lower bounds first and
the other two variants of the \textsc{BlindRandomBadRatio} (see \ref{fig:blind-rbr-variants}).
